\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)

### 2015-09-30 draft1leslieandkenny.Rnw
Made some progress on model and graphics.

require(xtable)
#require(foreign)
#require(arm)
require(tabplot)
require(dplyr)
require(car)
require(lme4)
require(ggplot2)
require(lattice)
require(effects)
@

@@

# Combine nitrogen and inoc/cntl
fert2$n.status <- with(fert2, interaction(n.trt, status))
@

@@

@@

<<interactionplot,echo=FALSE>>=
qplot(x = n.trt, y = infected, geom = 'boxplot', color = status,
      facets = .~variety, data = fert2) +
  theme(axis.text.x = element_text(angle = 90))
@

<<heatmap,echo=FALSE>>=
## Heatmap with meaningful ordering

# Order the dataset
#fert.ordered <- arrange(fert2, block, variety, status, n.trt)
fert.ordered <- arrange(fert2, variety, block, status, n.trt)

@@

# Plot the heatmap, with zeros in black and segments separating the blocks
par(mar = c(5, 10, 6, 2))
image(x = seq(0.5, 5.5, 1/6), y = 1:6, z = infected.arranged,
      col = c('black', heat.colors(30)), xlab = '', ylab = '',
      zlim = c(0, 30), yaxt = 'n', xaxt = 'n')
title('Infection Counts', line = 4)
segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = 'white')
axis(3, labels = rep(levels(fert2$block), 5),
     at = seq(7/12, 5 + 5/12, 1/6), cex.axis = 0.75)
axis(2, labels = levels(fert2$n.status), at = 1:6, las = 2)
axis(1, labels = levels(fert2$variety), at = 1:5)

# Legend
par(mar = c(5, 1, 6, 2))
image(y = -0.5:30.5, z = matrix(0:30, nrow = 1), axes = FALSE, ylab = '',
      col = c('#000000', heat.colors(30)), zlim = c(0, 30))
title('Legend', line = 1.5)
axis(4)
@

<<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
tableplot(fert.ordered, select = c(infected, variety, block, n.trt, status),
          nBins = 30, sortCol = variety)
@

The \verb|itableplot| function in the \verb|tabplot| package is another
useful tool for visualizing the data and identifying patterns.

\subsection{Convergeance Warning Messages}

<<models,cache=TRUE>>=
glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status +
                    (1|block/row),
                  control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert2)
summary(glmm3way)

glmm2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status +
                    n.trt*status + (1|block/row),
                    control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert2)
summary(glmm2way)

Anova(glmm3way)
Anova(glmm2way)
anova(glmm2way, glmm3way)
@

<<plots1,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert2)
plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert2)
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
plot(allEffects(glmm3way), multiline = TRUE, type = 'response',
     se=TRUE, ylim = c(0, 1), ci.style='bars', x.var = 'n.trt', rotx = 45,
     layout = c(5, 1))
@

@@

noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status +
                    (1|block/row), subset = variety != 'SNMS',
                    control = glmerControl(optimizer = 'bobyqa'),
                    family = binomial, data = fert2)
summary(noSNMS3way)
Anova(noSNMS3way)

noSNMS2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status +
                    n.trt*status + (1|block/row), subset = variety != 'SNMS',
                    control = glmerControl(optimizer = 'bobyqa'),
                    family = binomial, data = fert2)
summary(noSNMS2way)
Anova(noSNMS2way)

@@

<<plots3,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
plot(resid(noSNMS3way) ~ fitted(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

@@

>>>>

require(xtable)
#require(foreign)
#require(arm)
require(tabplot)
require(car)
require(lme4)
require(lattice)
@

@@

# Combine nitrogen and inoc/cntl (for the heatmap labels)
fert2$n.status <- with(fert2, interaction(n.trt, status))
@

<<fakedata,echo=FALSE>>=
# Create some fake data with no differences anywhere
set.seed(983724)
fert.sim1 <- data.frame(
  'block' = c(rep('A', 30), rep('B', 30), rep('C', 30), rep('D', 30), rep('E', 30), rep('F', 30)),
  'variety' = c(rep('Var1', 6), rep('Var2', 6), rep('Var3', 6), rep('Var4', 6), rep('Var5', 6)),
  'n.application' = rep(c('FALL', 'FALL', 'EARLY SPRING', 'EARLY SPRING', 'LATE SPRING', 'LATE SPRING'), 30),
  'status' = rep(c('CNTL', 'INOC'), 90),
  'total' = rep(30, 180),
  'infected' = rnbinom(180, 4, 7/9))
#  'infected' = sample(fert2$infected, 180, replace = TRUE))
@

<<fakepre,echo=FALSE>>=
# Create a row factor
fert.sim1$row <- with(fert.sim1, interaction(variety, block))

# Correctly order nitrogen application levels, the default is alphabetical
fert.sim1$n.trt <- relevel(fert.sim1$n.application, 'FALL')

# If status=INOC is the most interesting case, make INOC the baseline
fert.sim1$status2 <- relevel(fert.sim1$status, 'INOC')
@

@@

\noindent Because the response variable is a Binomial count, and because
the design includes random assignment of variety to rows within blocks,
a binomial generalized linear mixed model is appropriate. This model
will estimate the probability of infection for each
\verb|variety:treatment:status| combination while controlling for
differences between blocks and between rows in the same block.

\noindent The model is
$$
y_i\sim\mathrm{Binomial}(30, p_i),
$$
$$
\mathrm{logit}(p_i)=\mathbf{x}_i\mathbf{\beta}+b_{0,j[i]}+b_{1,j[i],k[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection,
\item \(\mathbf{x_i}\) is a row vector containing indicator variables for
the variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(\mathbf{\beta}\) is a column vector of fixed-effect coefficients, and
\item \(b_{0,j}\sim\mathrm{N}(0,\sigma^2_{b_0})\) and
\(b_{1,j,k}\sim\mathrm{N}(0,\sigma^2_{b_1})\) are random effects for block
and row, respectively.
\end{itemize}

\noindent The assumptions for this model are that plants within a plot
are independent and have the same probability of being infected, and
that plots within a row are endependent of each other.

\noindent Advantages of the Binomial GLMM include:
\begin{itemize}
\item It estimates and controls for variability between rows and blocks.
\item It avoids confounding variety effects with row effects and correctly
models the rows and nested within blocks.
\item Random effects can explain extra variation beyond what the
fixed-effect Binomial GLM can account for.
\end{itemize}

\noindent Disadvantages of this model include:
\begin{itemize}
\item It must be fit by numerically maximizing the likelihood with an
iterative algorithm. Including a large number of parameters (such as
third-order interaction coefficients) can cause the algorithm to fail
to converge.
\item The domain of the logit function is \((0, 1)\). If a large proportion
of the data are zeros, meaning an infection probability is near 0, the
algorithm will have difficulty estimating the coefficients. In this
situation, estimates may be unreliable or the algorithm may not converge.
\item We are not aware of any tools that can fit a Binomial GLMM with
overdispersion. If there are additional sources of variation that the
model does not account for, the standard error estimates will be too small.
\end{itemize}

@@

\noindent Before fitting a model, we recommend starting by creating plots
for exploratory data analysis. This is useful for seeing which interactions
may be present, and for identifying patterns that may represent violations
of assumptions or potential computational problems.

\noindent A good way to visually compare varieties, treatments, and
inoculation status is with boxplots of infection counts for each
combination of factor levels. The \verb|qplot| function in the
\verb|ggplot2| package is a simple way to do this.

\begin{figure}
\caption{Example interaction plot using \texttt{qplot}.}
<<interactionplot,echo=FALSE,message=FALSE>>=
require(ggplot2)
qplot(x = n.trt, y = infected, geom = 'boxplot', color = status,
      facets = .~variety, data = fert.sim1) +
  theme(axis.text.x = element_text(angle = 90))
@
\end{figure}

\noindent We noticed that infection counts of 0 were recored in 67 of the
180 plots in Nar's dataset. This is a large fraction of the data; if the
distribution of these zeros is related to the variables in the model then
it may be impossible to estimate coefficients for factor levels where
infection rates are nearly zero.

\noindent We recommend sorting the data and creating a heatmap to look
for patterns of zeros. Figure 2 shows an example using simulated data.
The most serious problems will come from the zeros, so we chose a contrasting
color (black) for plots with infection counts of 0.

\begin{figure}
\caption{Example heatmap visualization for identifying patterns of zeros.}
<<heatmap,echo=FALSE,message=FALSE>>=
## Heatmap with meaningful ordering

# Order the dataset, using the arrange() function from the dplyr package
require(dplyr)
fert.ordered <- arrange(fert.sim1, variety, block, status, n.trt)

@@

# Plot the heatmap, with zeros in black and segments separating the blocks
par(mar = c(5, 10, 6, 2)) # Set big margins
image(z = infected.arranged, y = 1:6,

      # Variety blocks are 1 unit wide, centered at 0.5, 1.5, etc
      x = seq(0.5, 5.5, 1/6),

      # Use black for 0, and use heatmap colors (red-orange-yellow-white) for 1 to 30
      col = c('black', heat.colors(30)), zlim = c(0, 30),
      xlab = '', ylab = '', yaxt = 'n', xaxt = 'n') # Don't automatically create axes or labels

# Use white line segments to visually separate the varieties
segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = 'white')

# Place a title at the top, and label Varieties, treatment:status levels, and blocks around the image
title('Infection Counts', line = 4)
axis(3, labels = rep(levels(fert.ordered$block), 5), cex.axis = 0.75,

     # Each block is plotted in a column with width 1/6, so put the labels in the middle
     at = seq(7/12, 5 + 5/12, 1/6))
axis(2, labels = levels(with(fert.ordered, interaction(n.trt, status))), at = 1:6, las = 2)
axis(1, labels = levels(fert.ordered$variety), at = 1:5)

# Legend
par(mar = c(5, 1, 6, 2))
image(y = seq(-0.5, 30.5, 1), z = matrix(0:30, nrow = 1), axes = FALSE, ylab = '',
      col = c('#000000', heat.colors(30)), zlim = c(0, 30))
title('Legend', line = 1.5)
axis(4)
@
\end{figure}

\noindent If information about the physical layout of the field is
available, it would be possible to construct a similar heatmap to look
for spatial patterns. If correlation between nearby plots is seen,
this should be included in the model.

\noindent Another option for examining the data graphically is the
\verb|itableplot| function in the \verb|tabplot| package. This is a
useful interactive tool for visualizing large datasets and identifying
patterns.

\begin{figure}
\caption{Example \texttt{itableplot} output.}
<<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
tableplot(fert.ordered, select = c(infected, variety, block, n.trt, status),
          nBins = 30, sortCol = variety)
@
\end{figure}

\subsection{Model Fitting and Possible Issues}
\noindent The
<<models1,warning=FALSE,results='hide',cache=TRUE>>=
glmm3way.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
                      control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
                      family = binomial, data = fert.sim1)
@

<<models2,warning=FALSE,results='hide',cache=TRUE>>=
noVar1.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
                      control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
                      family = binomial, data = fert.sim1, subset = variety!='Var1')
@

<<models3,warning=FALSE,results='hide',cache=TRUE>>=
glmm3way.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
                      control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
                      family = binomial, data = fert.sim1)
@

<<models4,echo=FALSE,warning=FALSE,results='hide',cache=TRUE>>=
glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
                  control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert2)
#summary(glmm3way)

glmm2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
                    n.trt*status2 + (1|block/row),
                    control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert2)
#summary(glmm2way)

#Anova(glmm3way)
#Anova(glmm2way)
#anova(glmm2way, glmm3way)
@

<<plots1,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert.sim1)
plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert.sim1)
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert.sim1$variety))
@

@@

noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
                    (1|block/row), subset = variety != 'SNMS',
                    control = glmerControl(optimizer = 'bobyqa'),
                    family = binomial, data = fert2)
summary(noSNMS3way)
Anova(noSNMS3way)

noSNMS2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
                    n.trt*status2 + (1|block/row), subset = variety != 'SNMS',
                    control = glmerControl(optimizer = 'bobyqa'),
                    family = binomial, data = fert2)
summary(noSNMS2way)
Anova(noSNMS2way)

@@

<<plots3,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
plot(resid(noSNMS3way) ~ fitted(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

\begin{figure}
\caption{The fitted model can be nicely summarized with plots from the \texttt{effects} package.}
<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
require(effects)
plot(allEffects(glmm3way.sim), multiline = TRUE, type = 'response', se = TRUE,
     ci.style='bars', x.var = 'n.trt', rotx = 45, layout = c(5, 1),
     main = 'Variety:Treatment:Inoculation Effects Plot',
     xlab = 'Nitrogen Application Date', ylab = 'Infection Probability',
     key.args = list(title = 'Inoculation Status'))
@
\end{figure}

@@

\section{Appendix: R Code}
\noindent Since Nar expressed a desire for assistance with R, we have
included the code we used to create the output for this report. Note
that we used a simulated dataset.

\subsection{Data Preprocessing}
\noindent We created variable with a unique level for each row, since row
it was not included in the dataset. We also changed the baseline levels
of \verb|n.application| and \verb|status|.
<<fakepre,eval=FALSE>>=
@

\subsection{\texttt{qplot}}
<<interactionplot,eval=FALSE>>=
@

\subsection{Heatmap}
Nar only needs to change \verb|fert.sim1| inside the \verb|arrange| call
to the name of his data frame. The rest of this code is self-contained and
will create the heatmap with appropriate labels for varieties, treatments,
and blocks.
<<heatmap,eval=FALSE>>=
@

\subsection{Summary Effects Plot}
<<plots2,eval=FALSE>>=
@
